\large{\bf Looking forwards and backwards through locally regulated populations }

### 2022-04-05 forwards-and-backwards.tex
edits on the plane

\newenvironment {proof}{{\noindent\bf Proof.}}{\hfill $\Box$ \medskip}

@@

\renewcommand{\theequation}{\arabic{section}.\arabic{equation}}
\def \non{{\nonumber}}
\def \hat{\widehat}
\def \tilde{\widetilde}
\def \bar{\overline}
\newcommand{\IP}{\mathbb P}
\newcommand{\IQ}{\mathbb Q}
\newcommand{\IE}{\mathbb E}
\newcommand{\IR}{\mathbb R}
\newcommand{\IZ}{\mathbb Z}
\newcommand{\IN}{\mathbb N}
\newcommand{\IT}{\mathbb T}
\newcommand{\IC}{\mathbb C}
\newcommand{\ind}{\mathbf{1}}

@@

\begin{document}
\pagenumbering{arabic}

@@

@@

The porous medium equation with logistic growth is an example of a
reaction-diffusion equation with a nonlinear diffusion term.
Such equations are widely used in a number of contexts in biology in which
it is clear that motility within a population varies with population density.
For example, density dependent dispersal is a common feature in spatial
models in ecology, eukaryotic cell biology, and avascular tumour growth;
see e.g.~\cite{sherratt:2010} and references therein for further discussion. 
We shall pay particular attention to the case in which the equation can be 
thought of as modelling the density of an expanding population. 
The specific equation that we use to illustrate the implications of
our results is one of a class used to study pattern formation in 
bacterial colonies. In particular, it has been suggested as a model for
the expansion of bacteria of the type {\em Paenbacillus dendritiformis} on
a thin layer of agar in a petri dish, 
\cite{cohen/golding/kozlovsky/benjacob/ron:1999}.

@@

@@

\plr{Description of the model:}

\plr{Main results (theorems without lookdowns):}

\comment{
There are two dichotomies:
is the limiting process deterministic or a generalized superprocess?
See Figure \ref{fig:super_vs_det_2d} for simulations.
And, are interactions local in the limit or not?
Below we have results for three of these cases;
as for the remaining one (local, superprocess)
TODO: write a note about what happens there (should be possible in 1D, not in higher dimensions).
}

We shall suppose that we start with a population of
$\mathcal{O}(N)$ individuals per unit area distributed in $\mathbb{R}^d$.
In applications we would typically take $\mathbb{R}^2$ or a subset thereof,
although our mathematical results apply in any spatial dimension.
We shall denote the state of the population at time $t$ by $\mathcal{S}_t$
and we shall abuse notation by treating $(\mathcal{S}_t)_{t \geq 0}$
as both the set of locations of the individuals
and a counting measure in which each individual has mass $1$, 
with the understanding that multiple 
points are treated as distinct individuals.  
The parameter $N$ will tend to infinity in our scaling limits, and the 
measure $(\overline{\eta}_t)_{t \geq 0} := (N^{-1} \mathcal{S}_t)_{t \geq 0} $ will converge (weakly) in the space of 
$\sigma$-finite measures on $\mathbb{R}^d$.

It will be simplest to think of the population as evolving in continuous time
and giving birth to one offspring at a time.
The rate at which a juvenile offspring is produced by an individual
will depend on the local population density.
We will write $\gamma(x,\overline{\eta})$ for the rate at which an
individual at location $x$ produces a juvenile offspring
when the state of the population is $\overline{\eta}$,
and $q(x,dy)$ for the distribution of the location of the juvenile 
when the parent is at $x$.
These offspring do not necessarily survive to be counted in the population:
we write $r(y, \overline{\eta})$ for the 
probability of successful establishment (survival to maturity)
of a juvenile at $y$.
It is assumed that maturity is achieved instantaneously,
and we shall only track mature individuals.
Finally, each mature individual located at the point $x$ dies at 
instantaneous rate $\mu(x,\overline{\eta})$.

\plr{I moved up the description of the interaction kernel here,
    so that the whole description of the model is in one place.
    I think this is good, but: some of our results don't depend on this assumption,
    which we should make more clear here. And: do we always use the heat kernel
    for $\varrho$? If so, we should just say so here.  }
Since we are interested in models where interactions are \emph{local},
i.e., reproduction, establishment, and death depend only
on the geographically local environment.
To this end,
we suppose that the birth rate of an individual at $x$
depends on the state of the population only through
a locally averaged density.
Concretely,
given an \emph{interaction distance} $\varepsilon$,
and a probability density $\varrho$ (the \emph{interaction kernel})
we will suppose that
$\gamma(x, \overline{\eta})$, $r(x, \overline{\eta})$, and $\mu(x, \overline{\eta})$
are functions only of $x$ and
\begin{align*}
    \varrho_\varepsilon * \overline{\eta}(x)
    := \int \varepsilon^{-d} \varrho(x/\varepsilon)(x-y) \overline \eta(dy)
    .
\end{align*}
To make the notation less complex, however,
we'll not formalize this until later (\S~\ref{sec: Main Theorems on Convergence}).

\begin{remark}
This model can be easily generalised into one
in which the number of offspring is drawn from a distribution,
\parencite[as in][]{etheridge/kurtz:2018}
and the main statements in our paper will still hold.
We have adopted this simple birth model
so we may demonstrate our proof in a more palatable manner.
\end{remark}

@@

\plr{Main theorem 1:}

\begin{theorem}\label{teo: Main Convergence Theorem - non-linear superprocess}
 Let $(\overline{\eta}^{\varepsilon, \theta,N}(t))_{t \geq 0}$ be the re-scaled model defined in Definition \ref{Def: Rescaled Models} and assume that as $\theta \to \infty$, $N \to \infty$, $\varepsilon(\theta)$ remains constant and $\theta/N \to \ell$.\\~\\
    The sequence of stochastic processes $(\overline{\eta}^{\varepsilon,\theta,N}_t)_{t \geq 0}$ converges weakly (analytically) in distribution as $\theta \to \infty$ to a measure-valued process $(\Xi^\varepsilon_t)_{t \geq 0}$ which satisfies that, for every $f:\mathbb{R}^d \rightarrow \mathbb{R}$,
\begin{align*} M_t &:= \langle f(x), \Xi^\varepsilon_t(dx) \rangle - \langle f(x), \Xi^\varepsilon_0(dx) \rangle \\ &- \int_0^t \big\langle \gamma(x,\varrho_{\varepsilon}*\Xi^{\epsilon}_{s}(x)) \mathcal{B}(f(\cdot)r(\cdot,\varrho_{\varepsilon}*\Xi^{\epsilon}_{s}(\cdot))(x) + f(x)  F(x,\varrho_{\varepsilon}*\Xi^{\epsilon}_{s}(x)), \Xi^\varepsilon_s(dx) \big\rangle ds, \end{align*}
is a martingale. The quadratic variation of $M_t$ is given by 
\begin{align*}
[ M ]_t = \int_0^t \ell\big\langle\gamma\big(x,\varrho_{\varepsilon}*\Xi^{\epsilon}_{s}(x)\big) r\big(x,\varrho_{\varepsilon}*\Xi^{\epsilon}_{s}(x)\big) f^2(x), \Xi^\varepsilon_s (dx) \big\rangle ds. 
\end{align*}
In other words, the limit $(\Xi^\varepsilon_t)_{t \geq 0}$ is a measure-valued process with non-local interactions.
Furthermore, if $\ell=0$ the limit is deterministic. In that case, if $\overline{\eta}^{\varepsilon,\theta}_0$ possesses a density in $L^2(\mathbb{R}^d)$, and the functions $\gamma$, $F$ are Lipschitz, then $(\overline{\eta}^{\varepsilon,\theta}_t)_{t \geq 0}$ possesses a density in $L^2(\mathbb{R}^d)$ and  $(\overline{\eta}^{\varepsilon,\theta}_t)_{t \geq 0}$ converges to a unique limit $(\Xi^\varepsilon(t))_{t \geq 0}$ with density $\varphi^\varepsilon(t,x)$ that solves (weakly) the mollified PDE of the form
\begin{equation}\label{Eq: non-local density PDE equation}
\partial_t \varphi(t,x) = r\big(x,\varrho_{\varepsilon}*\varphi(t,x)\big) \mathcal{B} \Big(\varphi(t,\cdot)\gamma\big(x,\varrho_{\varepsilon}*\varphi(t,\cdot)\big)\Big)(x) + \varphi(t,x) F\big(x,\varrho_{\varepsilon}*\varphi(t,x)\big)    
\end{equation}
\end{theorem}

\plr{Main theorem 2:}

\begin{theorem}\label{teo: Main Convergence Theorem - local pde limit}
 Assume that $\theta/ N \to 0 $ and assume that $\varepsilon(\theta) \to 0$. The sequence of mollifiers $\varrho_{\varepsilon}$ thus converges weakly to the Dirac delta mass, i.e. 
$\int_{\mathbb{R}^d} f(x)\varrho_{\varepsilon}(y-x)dx \to f(y)$. Assume furthermore that the following two conditions holds:
\begin{enumerate}
    \item The rate at which the local interaction strength $\varepsilon:=\varepsilon(\theta)$, speed $\theta$ and mass $\frac{1}{N(\theta)}$ scale satisfy the Equation 
\begin{equation} \label{condepsilon}
\frac{\theta}{N(\theta) \varepsilon(\theta)^{\frac{3 d p}{2}}} + \frac{1}{\theta \varepsilon(\theta)^{\frac{d p}{2}}} \xrightarrow[\theta \rightarrow \infty]{} 0,
\end{equation}

and $\overline{\eta}^{\varepsilon,\theta,N}_0$ is given by a Poisson point process with mean measure $\varphi_0(x)$.
    \item As $\varepsilon \to 0$, the sequence of functions $\{\varphi^\varepsilon(t,x): \varepsilon > 0 \}$ defined in Definition \ref{def: Weak solution formulation, mollified version} converges uniformly in space for any fixed time $t > 0$ to the function $\{\varphi(t,x): x \in \mathbb{R}^d, t >0\}$  which solves the partial differential equation 
    \begin{equation}\label{eq: local PDE limit generalised}
    \partial_t \varphi(t,x)=r(x,\varphi(t,x))\mathcal{B}\left(\varphi(t,\cdot) \gamma\big(\cdot,\varphi(t,\cdot)\big)\right)(x)+ \varphi(t,x)F\big(x,\varphi(t,x)\big)    
    \end{equation}
    with intial condition $\varphi(0,x)=\varphi_0(x)$.
\end{enumerate}
Then the sequence of stochastic processes $\{(\overline{\eta}^{\varepsilon, \theta,N}(t))_{t \geq 0}: \varepsilon > 0\}$ converges weakly (analytically) on finite dimensional distributions (in probability) to the deterministic process $\{\varphi(t,x): x \in \mathbb{R}^d, t >0\}$ defined in Equation \eqref{eq: local PDE limit generalised}.
\end{theorem}

\comment{Set-up and motivation for theorem on lineages.}
In \S \ref{Sec: Lookdown Constructions}, we will introduce a lookdown construction for the model described in \S \ref{sec: individual based model}. A lookdown construction assigns each particle a unique level which enables us to do a richer analysis. The main interest in using a lookdown construction for our
population processes is that it allows us to retain information 
about relatedness between individuals as we pass to the infinite
density limit. In particular, we are able to rigorously establish an equation for the motion of ancestral lineages in our limiting partial differential equation.

Consider a population evolving according to one of our deterministic
limiting models. We shall restrict ourselves
to the setting in which the population is expanding as a travelling wave
through $\IR$, with fixed profile $w(x)$ at rate $\mathfrak{c}$. That is 
$$\Xi_t(x)=w(x-\mathfrak{c}t),$$
for some fixed function $w$, and some $\mathfrak{c}\geq0$.  The profile $w$, if it exists, will satisfy
\begin{equation}
\label{profile}
r\Delta(\gamma w)+F(w)+\mathfrak{c}\nabla w=0.
\end{equation} 
It is convenient to work in a moving frame so that the wavefront is fixed. In this setting we want to investigate the ancestry line of an individual.\\
Note that in our model each individual has one parent only. As a result, as we go backwards in time, each individual has one unique ancestor at any time $t$.
\begin{definition}\label{def: Ancestry Line}
Let $w$ denote the wavefront of an evolving population that solves (\ref{profile}) and let $y \in supp(\Xi_t)$. We define $(L_s(y))_{s \geq 0}$, the ancestry line of an individual sampled at $y$, to be the position of the unique ancestor of $y$ at unit $s$ backwards in time.\\
We define $(\hat{T}_s)_{s \geq 0}$ to be the semi-group satisfying
$$\hat{T}_sf(y):=\mathbb{E}_{y}[f(L_s)]$$ and 
$\hat{\mathcal{L}}$ to be its corresponding generator. 
\end{definition}
The mathematical formulation for $L_s(y)$ will be made explicit once we introduce lookdown construction in \S \ref{Sec: Lookdown Constructions} as it allows us to trace lineages through stochastic differential equations, see \S \ref{sec: stochastic equations for lineages}.

\plr{Main Theorem 3}

\comment{TODO: clarify that this applies to either local or nonlocal population limits}

\begin{theorem} \label{teo:ancestral lineages}
    Let $w$ denote the wavefront of an evolving population that solves (\ref{profile}).
    Then the generator $\hat{\mathcal{L}}$ of the motion of an ancestral line of individuals in this population is given by
    \begin{equation}
    \label{eq: generator of lineage motions}
        \hat{\mathcal{L}} \phi
        =
        r\gamma\left(\Delta\phi+2\frac{\nabla(\gamma w)}{\gamma w}\nabla\phi\right)
        +\mathfrak{c} \nabla\phi.
    \end{equation}
\end{theorem}

\comment{TODO: short note on why it doesn't make sense to do this for a superprocess
         (it'd be Markov backwards in time given the population history probably,
         but it would be like a diffusion on a singular, evolving space or something)
}

>>>>

\newenvironment {proof}{{\noindent\bf Proof }}{\hfill $\Box$ \medskip}

@@

\renewcommand{\theequation}{\arabic{section}.\arabic{equation}}
\def \non{{\nonumber}}
\def \hat{\widehat}
\def \tilde{\widetilde}
\def \bar{\overline}
\newcommand{\IP}{\mathbb P}
\newcommand{\IQ}{\mathbb Q}
\newcommand{\IE}{\mathbb E}
\newcommand{\IR}{\mathbb R}
\newcommand{\IZ}{\mathbb Z}
\newcommand{\IN}{\mathbb N}
\newcommand{\IT}{\mathbb T}
\newcommand{\IC}{\mathbb C}
\newcommand{\ind}{\mathbf{1}}
\newcommand{\bigO}{\mathcal{O}}
\newcommand{\grad}{\nabla}

%%%%%%% notation
\newcommand{\DG}{\mathcal{B}}  % generator of the dispersal process
\newcommand{\meanq}{\vec b}    % mean of dispersal, times theta
\newcommand{\covq}{\Sigma}     % covariance matrix of dispersal, times theta
\newcommand{\kernel}{\varrho}  % interaction kernels
\newcommand{\smooth}[1]{\kernel_{#1} \! * \!}  % convolution by the interaction kernel
\newcommand{\wavespeed}{\mathfrak{c}}    % speed (vector) of a wave
\newcommand{\Lgen}{\mathcal{L}}    % generator of a lineage
\newcommand{\Pgen}{\mathcal{P}}    % generator of the population process

@@

\begin{document}

@@

\plr{I have not edited the introduction yet! It's just a pile of words. Skip to the next section.}

@@

@@

% % % % % % % % % % % % % %
\subsection{Structure of the paper}

In the paper we study scaling limits of spatial popultaions,
obtaining convergence of both the population process
(i.e., the population density as a function of time,
although strictly speaking it is a measure that may not have a density)
as well as of lineages traced back through such a population.
We retain linages through the rescaling process
by means of a lookdown construction, which adds technical difficulty
to an already difficult topic.
Furthermore, convergence results for the lookdown process
require convergence of the population process.
So, we do not introduce the lookdown process until Section X.

First, in section X, we describe the model and the main theorems (Theorems X and Y).
Next, in section X, we discuss a few striking consequences of these results,
namely, X Y and Z.
In section X, we provide heuristic explanations
of why the theorems ought to be true,
and some key ideas behind them.
In Section X we define and discuss the lookdown process that retains the levels.
This is followed by the technical proofs in section X.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Notes to ourselves}

\plr{For recording/discussing notation decisions.}

$X$ is the original, unscaled process,
and $\eta(t) = X(t\theta)/N$.

When the process has a density, it is $\varphi(t, x) = d \eta_t(dx)/dx$.

@@

The model we work with is a fairly general model
of individuals living in continuous space.
We shall suppose that we start with a population of
$\bigO(N)$ individuals per unit area distributed in $\mathbb{R}^d$.
For applications, $d=1$ or $d=2$, but our results apply more generally.
The population will be distributed over a bounded region,
so the total number of individuals will also be $\bigO(N)$.
We describe the state of the population at time $t$
with a counting measure $X(t)$,
which assigns one unit of mass to the location of each individual.
Locations need not be unique, so some locations may have a mass of greater than 1.

It will be simplest to think of the population as evolving in continuous time
and giving birth to one offspring at a time.
Population dynamics are controlled by three quantities,
birth, establishment, and death,
each of which can depend on spatial location and local population density.
Each individual gives birth at rate $\gamma$ to a single offspring,
and the offspring
then disperses according to a kernel $q(\cdot)$ away from the location of the parent.
We assume that $q$ is the density of a multivariate Gaussian,
allowing a nonzero mean and inhomogeneous variance.
The offspring does not necessarily survive to be counted in the population:
it ``establishes'' with probability $r$,
or else it dies immediately.
Independently, each individual dies with rate $\mu$.

We aim to obtain universal behavior by taking the density, $N$, to infinity,
and also scaling time by a factor of $\theta$,
so that defining $\eta(t) = X(\theta t) / N$,
the process $\{\eta(t)\}_{t \ge 0}$
will converge to a suitable measure-valued process
as $N$ and $\theta$ tend to infinity,
with the nature of the limit depending on how they tend to infinity together.
Demographic rates in our model
will be determined by local population density relative to $N$,
so that population dynamics depend more naturally on $\eta$ (rather than $X$).
However, to avoid confusion arising from the time scaling,
we describe the model in terms of $X$ next.
See Figure~\ref{fig:model_setup} for a conceptual depiction.

\begin{figure}
    \begin{center}
        FIGURE
    \end{center}
    \caption{
        Conceptual figure of the model.
        Left: (cartoon) plot of $\gamma$, $r$, $\mu$, and $F$ against density,
        showing the equilibrium point.
        Right: cartoon of interaction and dispersal distances.
        \label{fig:model_setup}
    }
\end{figure}

Birth, establishment, and death can depend on the location of the individual
and the local population density.
Since we would like the population density to scale with $N$,
these are functions of $X/N$, i.e.,
the counting measure with mass $1/N$ placed at the location of each individual.
To measure local population density, we convolve the population measure with
an ``interaction kernel'' -- for simplicity, we take these to be Gaussian
with various scales, and write
$$
    p_{\sigma^2}(x) = \frac{1}{(2 \pi \sigma^2)^{d/2}} e^{-\|x\|^2 / 2 \sigma^2} .
$$
Given a nonnegative function $\gamma(x, m) : \IR^d \times \IR_{\ge 0} \to \IR_{\ge 0}$,
the birth rate of an individual at location $x$ when the state of the population is $X$
is $\gamma(x, \smooth{\gamma} X(x) / N)$,
where $\smooth{\gamma} X$ is the convolution of $X$
with $p_{\epsilon_\gamma^2}$.
Similarly, the establishment probability of an offspring at location $x$ is
is $r(x, \smooth{r} X(x) / N)$,
where $r(x, m) : \IR^d \times \IR_{\ge 0} \to [0, 1]$
and again $\smooth{r} p_{\epsilon_r^2}$.
Rather than defining a function $\mu(x, m)$ for the death rate,
it turns out to be more convenient to parameterize in terms of the net reproductive rate,
which we denote by $F(x, m)$,
scaled by $1/\theta$
so that the population density changes over a time scale of $\theta$,
i.e., we have ``near-critical branching''.
Informally, we would like that
$$
    F = \theta (r \gamma - \mu) ,
$$
and to arrange this we define
the death rate of an individual at $x$ to be
\begin{align} \label{eqn:mu_defn}
    \mu
    =
    \max\left\{0, 
        r(x, \smooth{r} X(x) / N) \gamma(x, \smooth{\gamma} X(x) / N)
        - \frac{1}{\theta} F(x, \smooth{F} X(x) / N)
    \right\} ,
\end{align}
where again $F(x, m) : \IR^d \times \IR_{\ge 0} \to \IR_{\ge 0}$
and $\kernel_F = p_{\epsilon_F^2}$.

In summary, demographic parameters depend on local density measured
for birth over a scale of $\epsilon_\gamma$,
for establishment over a scale of $\epsilon_r$,
and for net reproductive rate over a scale of $\epsilon_F$.
Note that death rates depend in principle on population densities at three different scales,
so that we could write $\mu = \mu(x, \smooth{\gamma} X(x) / N, \smooth{r} X(x) / N, \smooth{F} X(x) / N)$.
It might seem like
we have introduced unnecessary complexity
with these three different spatial scales on which interactions occur.
However,
it will turn out to be useful in the proofs to have $\epsilon_r < \epsilon_\gamma$,
although in the limit they can be taken equal to each other,
so informally we think of local density as being determined on a scale of $\epsilon$.
The value of $\epsilon_F$ is unconstrained.

\begin{remark}
Although this model allows fairly general birth and death mechanisms,
there are a number of limitations.
Perhaps most obviously, to simplify the notation
individuals give birth to only one offspring at a time,
although this restriction could be easily lifted
\citep[as in][]{etheridge/kurtz:2018}.
Furthermore, individuals do not move during their lifetime,
and the age of an individual does not affect their fecundity or death rates.
Finally, there is no notion of mating,
so availability of mates will not restrict the spread of a population,
and the lineages we follow will be uniparental.
For these reasons, the model is most obviously applicable to bacterial populations,
although we do not anticipate that incorporation of these complications
will change the general picture.
\end{remark}

We next summarize the dynamics of the rescaled model $\eta(t) = X(\theta t)/N$,
and state some assumptions on the parameters.

\begin{definition}[Model parameterization and assumptions]
    \label{def:model_setup}
    For each $\theta \ge 0$,
    let $\left(\eta^\theta(t)\right)_{t \ge 0}$
    be the continuous-time measure-valued Markov process described above
    with birth rates, establishment probabilities, and death rates
    determined as follows.
    Suppose $\gamma(x, m)$, $r(x, m)$, and $F(x, m)$
    are bounded, uniformly continuous functions
    from $\IR^d \times \IR_{\ge 0}$ to $\IR_{\ge 0}$.
    Furthermore, suppose $r_\theta$ takes values in $[0, 1]$
    and has uniformly bounded second derivatives,
    and that $(x, m) \mapsto m^2 \gamma(x, m)$ is unformly bounded.
    Fix nonnegative values $\epsilon_r$, $\epsilon_\gamma$, and $\epsilon_F$,
    with $\epsilon_r < \epsilon_\gamma - 1/\theta$.
    Suppose $F: \IR^d \times \IR_{\ge 0} \to \IR$ is bounded and uniformly continous
    and define
    \begin{align}
        \mu(x, \eta)
        =
        r(x, \smooth{r} \eta(x))
        \gamma(\smooth{\gamma} \eta(x))
        - \frac{1}{\theta}
        F(x, \smooth{F} \eta(x)) .
    \end{align}
    Let $N$ be a positive number determined by $\theta$.
    When the state of the process is $\eta$,
    offspring of mass $1/N$ are produced at location $x$
    with rate $\theta \gamma(x, \smooth{\gamma} \eta(x)) N \eta(dx)$
    and adults die at location $x$ with rate $\theta \mu(x, \eta) N \eta(dx)$,
    (i.e., at rates $\theta \gamma$ and $\theta \mu$ per capita respectively).
    Offspring disperse to a location offset from the parent's location
    by an independent Gaussian with mean $\meanq/\theta$
    and covariance matrix $\covq / \theta$;
    an offspring at $x$ establishes instantaneously
    with probability $r(x, \smooth{r} \eta(x))$, or else dies.
\end{definition}

Since we are taking a diffusion limit,
we will have use for the following notation:

\begin{definition}[Dispersal generator]
    \label{def:dispersal_generator}
    Recall that we have defined the dispersal kernel,
    $q_\theta(x, dy)$,
    to be the density of a multivariate Normal
    with mean $\meanq/\theta$ and covariance matrix $\covq/\theta$.
    This implies that if we define, for $f : \IR^d \to \IR$,
    \begin{align}
    \DG f(x) = \sum_i \meanq_i \partial_{x_i} f(x)
        + \sum_{ij} \covq_{ij} \partial_{x_i} \partial_{x_j} f(x)
    \end{align}
    then
    \begin{align}
        \theta \int \left(
            f(y) - f(x)
        \right) q_\theta(x, dy)
    \to \DG f(x) 
        \qquad \text{as } \theta \to \infty .
    \end{align}
\end{definition}

\paragraph{Remark:}
An equivalent way to describe the model
would be to say that an individual at $x$ gives birth at rate
$$
    \gamma(x, \smooth{\gamma} \eta(x))
    \int r(x, \smooth{r} \eta(y)) q(x, dy) ,
$$
and that offspring disperse according to the kernel
$$
    q_\theta^\mathfrak{m}(x, dy)
    :=
    \frac{
        r(y, \smooth{r} \eta(y)) q_\theta(x, dy)
    }{
        \int r(x, \smooth{r} \eta(z)) q_\theta(x, dz)
    } .
$$
Clearly, the random walk driven by this dispersal kernel
should be biased towards regions of higher establishment probability.
For comparison to future results,
it is interesting to write down the limiting generator:
\begin{align*}
    \lim_{\theta \to \infty}
    \theta \int (f(y) - f(x)) q_\theta^\mathfrak{m}(x, dy)
    &=
    \frac{
        \DG\left[ f(\cdot) r(\cdot, \smooth{r} \eta(\cdot)) \right](x)
        - 
        f(\cdot) \DG\left[ r(\cdot, \smooth{r} \eta(\cdot)) \right](x)
    }{
        r(x, \smooth{r} \eta(x))
    } .
\end{align*}
In the simplest case of unbiased, isotropic dispersal,
$\meanq = 0$ and $\covq = I$, so $\DG = \Delta$,
and this is equal to
\begin{align*}
    \Delta f(x) + 2 \grad f(x) \cdot \grad \log r(\cdot, \smooth{r} \eta(\cdot))(x) .
\end{align*}
However, we will see that following a lineage back through the population
has a different behavior yet again.

% % % % % % % % % % % % % % %
\subsection{Main results on population processes}

\comment{
Set-up: there are two dichotomies:
is the limiting process deterministic or a generalized superprocess?
See Figure \ref{fig:super_vs_det_2d} for simulations.
And, are interactions local in the limit or not?
Below we have results for three of these cases;
as for the remaining one (local, superprocess),
it should be possible in 1D, not in higher dimensions (brief note).
}

@@

\begin{theorem} \label{thm:nonlocal_convergence}
    Let $(\eta^\theta_t)_{t \geq 0}$
    be as defined in Definition \ref{def:model_setup}
    and assume that as $\theta \to \infty$, $N \to \infty$
    in such a way that $\theta/N \to \delta$.
    (However, interaction distances $\epsilon_r$, $\epsilon_\gamma$, and $\epsilon_F$
    remain constant.)

    The sequence of stochastic processes
    $(\eta^\theta_t)_{t \geq 0}$
    converges along subsequences in distribution as $\theta \to \infty$
    to a measure-valued process $(\eta_t)_{t \geq 0}$
    which satisfies that, for every $f:\mathbb{R}^d \rightarrow \mathbb{R}$,
    \begin{align} \label{eqn:mgale_problem}
        M_t
        &:=
        \langle f(x), \eta_t(dx) \rangle
        -
        \langle f(x), \eta_0(dx) \rangle
        \\ & \qquad
        -
        \int_0^t \big\langle
            \gamma(x, \smooth{\gamma} \eta_s(x))
            \mathcal{B}\left(
                f(\cdot) r(\cdot, \smooth{r} \eta_s(\cdot))
            \right)(x)
            +
            f(x)
            F(x, \smooth{F} \eta_s(x)),
            \eta_s(dx)
        \big\rangle ds,
    \end{align}
    is a martingale.
    The quadratic variation of $M_t$ is given by 
    \begin{align} \label{eqn:mgale_variation}
        [ M ]_t
        =
        \delta
        \int_0^t
        \big\langle
            \gamma\left( x, \smooth{\gamma} \eta_s(x) \right)
            r\left(x, \smooth{r} \eta_s(x) \right)
            f^2(x),
            \eta_s (dx)
        \big\rangle ds. 
    \end{align}
    In other words, the limit is a measure-valued process with non-local interactions.
    Furthermore, if $\delta = 0$ the limit is deterministic.
\end{theorem}

Theorem \ref{thm:nonlocal_convergence} provided convergence along subsequences;
i.e., for any sequence of values of $\theta$
there is a subsequence that converges.
However, if the limiting process is unique then the convergence is unqualified.
In the deterministic case, we have convergence to a nonlocal PDE.
\comment{Maybe the following two corollaries should be combined?}

\comment{
    \textbf{Question:}
    In our outline
    we said this corollary would be about uniqueness,
    but is it also about continuity of the limit (i.e., existence of a density)?
    A lot of my uncertainty in how to phrase these results
    comes from being unsure what exactly we can say about densities.
}

\begin{corollary}[Deterministic limits]
    \label{cor:nonlocal_pde_limits}
    In the context of Theorem~\ref{thm:nonlocal_convergence},
    if the limit is deterministic (i.e., $\theta / N \to \delta = 0$), $\eta_0$ has a density,
    and \comment{additional conditions? can we say something like this??}
    then any limit $\eta_t(dx)$ has a density $\varphi(t, x)$,
    which is a weak solution to the nonlocal partial differential equation
    \begin{equation} \label{eqn:nonlocal_pde}
        \partial_t \varphi(t, x)
        =
        r\left(x, \smooth{r} \varphi(t,x) \right)
        \mathcal{B} \left[
            \varphi(t, \cdot)
            \gamma\big( x, \smooth{\gamma} \varphi(t, \cdot) \big)
        \right](x)
        +
        \varphi(t, x)
        F\left(x, \smooth{F} \varphi(t,x) \right)
        .
    \end{equation}
    If equation \eqref{eqn:nonlocal_pde} has a unique solution,
    then $(\eta^\theta_t)_{t \ge 0}$ converges to that solution.

    In particular, if
    \comment{TODO: find some conditions under which this is true?}
    then we have convergence to a unique solution.
\end{corollary}

Informally, equation \eqref{eqn:nonlocal_pde} is the PDE
\begin{align} \label{eqn:pde}
    \dot \varphi = r \DG\left( \gamma \varphi \right) + \varphi F ,
\end{align}
where $\DG$ is an elliptic second order differential operator,
and the coefficients $r$, $\gamma$, and $F$
at $x$ depend nonlocally on $\varphi$
(i.e., through a smoothed version of $\varphi$).

\begin{corollary}[Superprocess limits] \label{cor:superprocess_uniqueness}
    In the context of Theorem~\ref{thm:nonlocal_convergence},
    if the martingale problem
    defined by equations \eqref{eqn:mgale_problem} and \eqref{eqn:mgale_variation}
    has a unique solution,
    then $(\eta^\theta(t))_{t \ge 0}$ converges to that solution
    as $\theta \to \infty$.
    \comment{
        Is this what we want to say here? Could be said better.
        TODO: see what Dawson \& Li say about uniqueness
        so we have at least some conditions under which this happens.
    }
\end{corollary}

Theorem~\ref{thm:nonlocal_convergence} shows convergence to the solution to a nonlocal PDE
of the form \eqref{eqn:nonlocal_pde}.
Since the coefficients at $x$ in \eqref{eqn:nonlocal_pde}
depend only on the region nearby to $x$,
we would often expect solutions of the nonlocal PDE to be close to the corresponding local PDE
as the interaction distances $\epsilon_r$, $\epsilon_\gamma$ and $\epsilon_F$
go to zero.
The following propositions give two concrete situations in which that is true.

\comment{I've not got straight exactly how we're framing this
    (namely, does $\eta$ have a density?).
    Also, is this one or two propositions --
    we said initially this would be two,
    one for FKPP and one for PME,
    but then maybe the argument was more general?
}

\begin{proposition}
    \label{prop:nonlocal_to_local}
    Suppose that $\eta_\epsilon(t,dx)$ is a solution to equation \eqref{eqn:nonlocal_pde}
    with $r(x, m) = 1$ for all $x$ and $m$
    and some $\epsilon_r < \epsilon_\gamma = \epsilon_F = \epsilon$
    \comment{and assumptions on $\gamma$? and $F$?}.
    Then $\eta_\epsilon(t, dx)$ converges \comment{in what sense?} as $\epsilon \to 0$
    to $u(t, x) dx$, which solves
    \begin{align} \label{eqn:PDE}
        \partial_t u(t, x)
        =
        \DG \left( \gamma(\cdot, u(t, \cdot) u(t, \cdot)  \right)(x)
        + F(x, u(t, x)) u(t, x) .
    \end{align}
\end{proposition}

Theorem \ref{thm:nonlocal_convergence} combined with Proposition \ref{prop:nonlocal_to_local}
imply that we can take the $\theta \to \infty$ limit
followed by $\epsilon \to 0$
to obtain solutions to the PDE \eqref{eqn:PDE}.
However, it is of substantial interest to know whether
we can take those two limit simultaneously.
The general case seems difficult,
but we can prove such ``diagonal'' convergence in the following situation.

\begin{theorem}[Convergence to a PDE]
    \label{thm:local_convergence}
    Let $(\eta^\theta_t)_{t \geq 0}$
    be as defined in Definition \ref{def:model_setup}
    and assume that as $\theta \to \infty$, $N \to \infty$
    and $\epsilon_r < \epsilon_\gamma = \epsilon_F = \epsilon \to 0$
    in such a way that $\theta/N \to 0$
    and
    \comment{TODO: some conditions on $\epsilon$ and $r$ and $\gamma$ and $F$}.
    Then the sequence of stochastic processes $(\eta^\theta_t)_{t \ge 0}$
    converges \comment{in some sense}
    to a measure-valued process with a density $\varphi(t, x)$
    that solves
    \begin{align}
        \partial_t \varphi(t, x)
        &=
        \text{whatever the final form is here} .
    \end{align}
\end{theorem}

% % % % % % % % % % % %
\subsection{Main results on lineages}

Now that we have established what we can say about how population density changes with time,
we turn to results on ancestral lineages,
i.e., how genealogical ancestry moves across the landscape.
Informally,
a \emph{lineage} $(L_t)_{t \ge 0}$
begun at spatial location $L_0 = x$
can be obtained by picking a focal individual uniformly at $x$,
and then for each amount of time $t$ in the past,
setting $L_t$ to the spatial location of the individual from whom
the focal individual inherits.
Since in our model individuals have only one parent, this is unambiguous.
Although we did not explicitly retain such information,
it is clear that
for finite $\theta$ (and hence, finite $N$),
one could construct the distribution of a lineage $(L_t)_{t=0}^T$
given the history of the population $(\eta^\theta_t)_{t = 0}^T$,
for each starting location to which $\eta^\theta_T$ assigns positive mass.
It is less clear, however, how to formally retain such information in the limit.
The \emph{lookdown construction} in Section~\ref{sec:lookdown}
provides just such a construction,
where lineages are retained in the $N \to \infty$ limit.
Roughly speaking,
this works by assigning to each particle a unique ``level''
that functions as a label and thus allows reconstruction of lineages,
but also (in some sense) orders individuals by eventual reproductive output,
allowing a formal limit.
See~\ref{etheridge/kurtz:2018} for an introduction to these ideas.
